---
title: "Linguistic Intergroup Bias in Wikipedia"
author: "David Garcia"
date: "February 25, 2016"
output:
  html_document:
    highlight: pygments
    keep_md: true
---

## Composing the Wikipedia Abstracts dataset

1. Abstracts and genders from Wikipedia.
Women Through the Glass Ceiling: Gender Asymmetries in Wikipedia. Claudia Wagner, Eduardo Graells-Garrido, David Garcia

https://github.com/clauwag/WikipediaGenderInequality/blob/master/data/person_text_1.csv.gz

```{r, engine='python', cache=TRUE, eval=FALSE}
import urllib
urllib.urlretrieve ("https://github.com/clauwag/WikipediaGenderInequality/blob/master/data/person_text_1.csv.gz?raw=true", "person_text_1.csv.gz")
urllib.urlretrieve ("https://github.com/clauwag/WikipediaGenderInequality/blob/master/data/person_text_2.csv.gz?raw=true", "person_text_2.csv.gz")
urllib.urlretrieve ("https://github.com/clauwag/WikipediaGenderInequality/blob/master/data/person_text_3.csv.gz?raw=true", "person_text_3.csv.gz")
urllib.urlretrieve ("https://github.com/clauwag/WikipediaGenderInequality/blob/master/data/person_text_4.csv.gz?raw=true", "person_text_4.csv.gz")
```

2. MPQA subjectivity lexicon
Theresa Wilson, Janyce Wiebe, and Paul Hoffmann (2005). Recognizing Contextual Polarity in Phrase-Level Sentiment Analysis. Proc. of HLT-EMNLP-2005. 
Distributed under GPL: http://www.gnu.org/licenses/gpl.html
http://mpqa.cs.pitt.edu/lexicons/subj_lexicon/

```{r, engine='python', cache=TRUE, eval=FALSE}
import zipfile
import urllib
#urllib.urlretrieve ("http://mpqa.cs.pitt.edu/data/subjectivity_clues_hltemnlp05.zip", "subjectivity_clues_hltemnlp05.zip")
with zipfile.ZipFile("subjectivity_clues_hltemnlp05.zip", "r") as z:
    z.extractall("subjectivity_clues")
```
                    
                    
```{r, engine='python', cache=TRUE, eval=FALSE}
import csv
import gzip
import nltk
from nltk.tag import pos_tag, map_tag

#load subjectivity dictionary
fin = open("subjectivity_clues/subjectivity_clues_hltemnlp05/subjclueslen1-HLTEMNLP05.tff","rt")
subjdic = dict()
for line in fin:
  fields = line.split(" ")
  contdic = dict()
  for field in fields:
    parts = field.split("=")
    if len(parts) == 2:
      contdic[parts[0]] = parts[1]
  w = contdic["word1"]
  subjdic[w] = contdic["type"], contdic["pos1"], contdic["priorpolarity"].replace("\n", "")
fin.close()

fout = open("textData.dat", "w")
fout.write("class" + "\t" + "year" + "\t"+ "gender" + "\t" + "nwords" + "\t" + "nadj" + "\t" + "npos" + "\t" + "nposadj" + "\t" + "nneg" + "\t" + "nnegadj" + "\n") 

for i in range(1,5):
  with gzip.open("person_text_"+str(i)+".csv.gz") as f:
    csvreader = csv.reader(f, delimiter=',', quotechar='"')
    for fields in csvreader:
      gender = fields[3]
      if fields[3] != "gender":
        jobclass = fields[2].replace("http://dbpedia.org/ontology/", "")
        year = fields[7]
        text = fields[12].decode("utf8").replace("\n", "").replace('"', '')

        texttokens = nltk.word_tokenize(text) #Tokenize into words
        posTagged = pos_tag(texttokens)
        simplifiedTags = [(word, map_tag('en-ptb', 'universal', tag)) for word, tag in posTagged] #Parts-Of-Speech tagging to detect adjectives
        nadj = nwords =  0
        npos = nneg = 0
        nposadj = nnegadj = 0

        for tag in simplifiedTags :
          if tag[1] != ".":
            nwords +=1
            word = tag[0].lower()
            
            sv = subjdic.get(word, -1)
            polarity = ""
            if sv != -1:
              polarity = sv[2]
            
            if tag[1] == "ADJ":
              nadj +=1
            if polarity == "positive":
              npos +=1
              if tag[1] == "ADJ":
                nposadj +=1
            if polarity == "negative":
              nneg +=1
              if tag[1] == "ADJ":
                nnegadj +=1
            if polarity == "both":
              npos +=1
              nneg +=1
              if tag[1] == "ADJ":
                nposadj +=1
                nnegadj +=1

        fout.write(jobclass  + "\t" + year  + "\t" +  gender + "\t" + str(nwords) + "\t" + str(nadj) + "\t" + str(npos) + "\t" + str(nposadj) + "\t" + str(nneg) + "\t" + str(nnegadj) + "\n") 

```


# Chi-squared tests of general proportions

```{r Stats, cache=TRUE, warning=FALSE}
library(xtable)
library(texreg)
df <- read.table("textData.dat", sep="\t", header=TRUE)
attach(df)

prop.test( c(sum(nposadj[gender=="male"]),sum(nposadj[gender=="female"])), c(sum(npos[gender=="male"]), sum(npos[gender=="female"])), alternative="greater")

# Effect size calculation
p0 <- sum(nposadj)/sum(npos)
p1 <- sum(nposadj[gender=="male"])/sum(npos[gender=="male"])
p2 <- sum(nposadj[gender=="female"])/sum(npos[gender=="female"])
print(paste("Cohen's W:", sqrt(((p0-p1)^2 + (p0-p2)^2)/p0)))
print(paste("% Change:", (p1-p2)/p1 * 100))

prop.test( c(sum(nnegadj[gender=="male"]),sum(nnegadj[gender=="female"])), c(sum(nneg[gender=="male"]), sum(nneg[gender=="female"])), alternative="less")

# Effect size calculation
p0 <- sum(nnegadj)/sum(nneg)
p1 <- sum(nnegadj[gender=="male"])/sum(nneg[gender=="male"])
p2 <- sum(nnegadj[gender=="female"])/sum(nneg[gender=="female"])
print(paste("Cohen's W:", sqrt(((p0-p1)^2 + (p0-p2)^2)/p0)))
print(paste("% Change:", (p1-p2)/p1 * 100))

```


# Does abstraction depend on the gender of the biography?

```{r, cache=TRUE, warning=FALSE, results='asis', fig.height=4, cache=TRUE}
library(xtable)
library(texreg)
df <- read.table("textData.dat", sep="\t", header=TRUE)
attach(df)
minW <- 50
f <- npos>0 & !is.na(year) & nwords >=minW & nposadj != npos & nposadj >0
posabs <- (nposadj/npos)[f]
posabsL <- log(posabs/(1-posabs))

par(mfrow=c(1,2))
hist(posabs, main="Positive adjective ratio")
hist(posabsL, main="Logit of Positive adjective ratio")
```

Logit models for the ratios:

$logit(r_+) = a_+ + b_+*\delta_F$ 

$logit(r_-) = a_- + b_-*\delta_F$ 

(we expect $b_+<0$ and $b_- >0$)


```{r Stats2, cache=TRUE, warning=FALSE, results='asis'}
G <- rep(0,sum(f))
G[gender[f] == "female"] <- 1
linModel <- glm(posabsL~G)

htmlreg(list(linModel), digits=4, single.row=TRUE, bold=0.05)

```


```{r, cache=TRUE, warning=FALSE, results='asis', fig.height=4}

f <- nneg>0 & !is.na(year) & nwords >=minW & nnegadj != nneg & nnegadj >0
negabs <- (nnegadj/nneg)[f]
negabsL <- log(negabs/(1-negabs))

par(mfrow=c(1,2))
hist(negabs, main="Negative adjective ratio")
hist(negabsL, main="Logit of Negative adjective ratio")
```


```{r , cache=TRUE, warning=FALSE, results='asis'}
G <- rep(0,sum(f))
G[gender[f] == "female"] <- 1
linModel <- glm(negabsL~G)

htmlreg(list(linModel), digits=4, single.row=TRUE, bold=0.05)

```



# Do these effects depend on the year when the person was born?

```{r , cache=TRUE, warning=FALSE}
df <- read.table("textData.dat", sep="\t", header=TRUE)
attach(df)
f <- npos>0 & !is.na(year) & nwords >=minW & nposadj != npos & nposadj >0
posabsL <- log(posabs/(1-posabs))
y <- year[f]
cor.test(posabsL, y)
G <- rep(0,sum(f))
G[gender[f] == "female"] <- 1
posdf <- data.frame(y=y, G=G, posabsL=posabsL)
yearModelPos <- glm(posabsL~G*y, data=posdf)
```

```{r , cache=TRUE, warning=FALSE}
f <- nneg>0 & !is.na(year) & nwords >=minW & nnegadj != nneg & nnegadj >0
negabsL <- log(negabs/(1-negabs))
y <- year[f]
G <- rep(0,sum(f))
G[gender[f] == "female"] <- 1
cor.test(negabsL, y)
negdf <- data.frame(y=y, G=G, negabsL=negabsL)
yearModelNeg <- glm(negabsL~G*y, data=negdf)
```

```{r , cache=TRUE, warning=FALSE, results='asis'}
htmlreg(list(yearModelPos, yearModelNeg), digits=4, single.row=TRUE, bold=0.05)
```

We visualize the interactions in the model

```{r , cache=TRUE, warning=FALSE, results='asis', eval=TRUE}
ys <- seq(min(posdf$y), max(posdf$y))

predsMale0 <- predict(yearModelNeg, newdata= data.frame(y=ys, G=rep(0,length(ys))), se.fit=TRUE)
#Preditions for male biographies in Negative model
predsMale <- exp(predsMale0$fit)/(exp(predsMale0$fit)+1)
predsMaleLow <- exp(predsMale0$fit-predsMale0$se.fit)/(exp(predsMale0$fit-predsMale0$se.fit)+1) 
predsMaleHigh <- exp(predsMale0$fit+predsMale0$se.fit)/(exp(predsMale0$fit+predsMale0$se.fit)+1) 

predsFemale0 <- predict(yearModelNeg, newdata= data.frame(y=ys, G=rep(1,length(ys))), se.fit=TRUE)
#Preditions for female biographies in Negative model
predsFemale <- exp(predsFemale0$fit)/(exp(predsFemale0$fit)+1)
predsFemaleLow <- exp(predsFemale0$fit-predsFemale0$se.fit)/(exp(predsFemale0$fit-predsFemale0$se.fit)+1) 
predsFemaleHigh <- exp(predsFemale0$fit+predsFemale0$se.fit)/(exp(predsFemale0$fit+predsFemale0$se.fit)+1) 

plot(ys, predsMale, col="red", type="l", ylim=range(c(predsMaleHigh, predsMaleLow, predsFemaleHigh, predsFemaleLow)), xlab="Birth year", ylab="Predicted r-")
lines(ys, predsMaleLow, col="red", lty=2)
lines(ys, predsMaleHigh, col="red", lty=2)
lines(ys, predsFemale, col="blue")
lines(ys, predsFemaleLow, col="blue", lty=2)
lines(ys, predsFemaleHigh, col="blue", lty=2)
legend(c("Male", "Female"), x="topright", col=c("red", "blue"), lwd=c(1,1)) 
```


```{r , cache=TRUE, warning=FALSE, results='asis', eval=TRUE}

predsMale0 <- predict(yearModelPos, newdata= data.frame(y=ys, G=rep(0,length(ys))), se.fit=TRUE)
#Preditions for male biographies in Positive model
predsMale <- exp(predsMale0$fit)/(exp(predsMale0$fit)+1)
predsMaleLow <- exp(predsMale0$fit-predsMale0$se.fit)/(exp(predsMale0$fit-predsMale0$se.fit)+1) 
predsMaleHigh <- exp(predsMale0$fit+predsMale0$se.fit)/(exp(predsMale0$fit+predsMale0$se.fit)+1) 

predsFemale0 <- predict(yearModelPos, newdata= data.frame(y=ys, G=rep(1,length(ys))), se.fit=TRUE)
#Preditions for female biographies in Positive model
predsFemale <- exp(predsFemale0$fit)/(exp(predsFemale0$fit)+1)
predsFemaleLow <- exp(predsFemale0$fit-predsFemale0$se.fit)/(exp(predsFemale0$fit-predsFemale0$se.fit)+1) 
predsFemaleHigh <- exp(predsFemale0$fit+predsFemale0$se.fit)/(exp(predsFemale0$fit+predsFemale0$se.fit)+1) 

plot(ys, predsMale, col="red", type="l", ylim=range(c(predsMaleHigh, predsMaleLow, predsFemaleHigh, predsFemaleLow)), xlab="Birth year", ylab="Predicted r+")
lines(ys, predsMaleLow, col="red", lty=2)
lines(ys, predsMaleHigh, col="red", lty=2)
lines(ys, predsFemale, col="blue")
lines(ys, predsFemaleLow, col="blue", lty=2)
lines(ys, predsFemaleHigh, col="blue", lty=2)
legend(c("Male", "Female"), x="topright", col=c("red", "blue"), lwd=c(1,1)) 
```

```{r , cache=TRUE, warning=FALSE, results='asis', eval=TRUE}
hist(posdf$y, main="Biographies per birth year", xlab="Birth Year")
```


